---
title: "Protein A vs Control Protein B with treatment 1, 2 and 3"
output: html_notebook
---

With this analysis I will be answering the following questions:
1. Does treatment 1, 2 and 3 have a significant effect on each protein's measured ratio?
2. Does protein A and Control protein B follow the same pattern for calibrated ratios?

### 1A. Installing packages

```{r message=FALSE}
# Installing packages
if(!require("dplyr")) {install.packages("dplyr")}
if(!require("readr")) {install.packages("readr")}
if(!require("ggplot2")) {install.packages("ggplot2")}
if(!require("tidyverse")) {install.packages("tidyverse")}
if(!require("patchwork")) {install.packages("patchwork")} # placing multiple ggplot net to each other
``` 

### 2A. Loading libraries & data

```{r message=FALSE}
# Loading libraries
library(dplyr)
library(readr)
library(ggplot2)
library(tidyverse)
library(patchwork)
```

```{r echo = FALSE}
ratio_df <- read_csv("proteinAvsB_ratios.csv",show_col_types=FALSE)
```

### 4A. Explore data structure, view & summary

```{r}
View(ratio_df) #Row length differs due to experimental results
str(ratio_df) # all data is numeric
summary(ratio_df)
str(ratio_df)
```

### 5A. changing column "pH"treatment" to class factor instead of numeric
```{r}
ratio_df$treatment <- as.factor(ratio_df$treatment)
levels(ratio_df$treatment)

```
### 6A. Visualize ratio distribution for each protein for treatment 1, 2 and 3.

```{r echo=FALSE}
violin_all <- ggplot(ratio_df,aes(x=treatment,y=ratio)) +
            geom_violin(aes(fill = treatment)) +
            geom_boxplot(outlier.color="red", alpha=0.5) +
            stat_summary(fun=median, geom="point", size=2, color="red") +
            labs(title = "Ratio for protein A and protein B") +
            ylim(0,50) +
            facet_wrap(~protein)
violin_all
```

### 7A. Organize data for removing outliers
Does protein A with treatment 1, 2 and 3 have significant outliers? 

Checking the statistics of the cleaned data:
```{r}
group_by(ratio_df, treatment, protein) %>%
  summarise(
    count = n(),
    mean = mean(ratio),
    sd = sd(ratio),
    median = median(ratio),
    IQR = IQR(ratio)
  )
```

```{r}
#--------Protein A with treatment 1---------
pAt1 <- ratio_df %>% # 736 observations
          group_by(treatment,protein,ratio) %>%
          filter(protein == "A" &
                 treatment == 1)
quantile1_pAt1 <- quantile(pAt1$ratio,0.25) # 17.37
quantile3_pAt1 <- quantile(pAt1$ratio,0.75) # 22.815
IQR_pAt1 <- quantile3_pAt1 - quantile1_pAt1 # 5.445
pAt1 <- pAt1 %>% 
          mutate(is_outlier = ratio < quantile1_pAt1 - (IQR_pAt1*1.5) |
                   ratio >  quantile3_pAt1 + (IQR_pAt1*1.5)) %>% 
          filter(!is_outlier) # remains 735 observations, 1 outlier removed

#--------Protein A with treatment 2---------
pAt2 <- ratio_df %>% # 176 observations
          group_by(treatment,protein,ratio) %>%
          filter(protein == "A",
                 treatment == 2)
quantile1_pAt2 <- quantile(pAt2$ratio,0.25) # 6.8
quantile3_pAt2 <- quantile(pAt2$ratio,0.75) # 9.955
IQR_pAt2 <- quantile3_pAt2 - quantile1_pAt2 # 3.155
pAt2 <- pAt2 %>% 
          mutate(is_outlier = ratio <  quantile1_pAt2 - (IQR_pAt2*1.5) |
                   ratio >  quantile3_pAt2 + (IQR_pAt2*1.5)) %>% 
          filter(!is_outlier) # remains 173 observations, 3 outliers removed


#--------Protein A with treatment 3---------
pAt3 <- ratio_df %>% # start with 316 observations
          group_by(treatment,protein,ratio) %>%
          filter(protein == "A",
                 treatment == 3)
quantile1_pAt3 <- quantile(pAt3$ratio,0.25) # 23.815
quantile3_pAt3 <- quantile(pAt3$ratio,0.75) # 31.4675
IQR_pAt3 <- quantile3_pAt3 - quantile1_pAt3 # 7.6525
pAt3 <- pAt3 %>% 
          mutate(is_outlier = ratio <  quantile1_pAt3 - (IQR_pAt3*1.5) |
                   ratio >  quantile3_pAt3 + (IQR_pAt3*1.5)) %>% 
          filter(!is_outlier) # remains 311 observations, 5 outliers removed

```

### 8A. View them in a histogram to check for distribution

```{r}
set.seed(0)
# Protein A treatment 1
violinplot_clean_pAt1 <- pAt1 %>% 
  ggplot(aes(x = "Treatment 1", y = ratio)) +
  geom_violin(alpha=0.5,fill = "red") + geom_boxplot(outlier.color="red", alpha=0.5) +
  xlab("")
  
violinplot_clean_pAt2 <- pAt2 %>% 
  ggplot(aes(x = "treatment 2", y = ratio)) +
  geom_violin(alpha=0.5,fill = "green") + geom_boxplot(outlier.color="red", alpha=0.5) +
  xlab("")
  

violinplot_clean_pAt3 <- pAt3 %>% 
  ggplot(aes(x = "treatment 3", y = ratio)) +
  geom_violin(alpha=0.5,fill = "blue") + geom_boxplot(outlier.color="red", alpha=0.5) +
  xlab("")
  
#--------------------------------------------------------------------
set.seed(5)
hist_pAt1 <- ggplot(pAt1,aes(ratio)) +
  geom_histogram(fill = "red", alpha = 0.5, binwidth = 1) +
  labs(title = "protein A, treatment 1")

hist_pAt2 <- ggplot(pAt2,aes(ratio)) +
  geom_histogram(fill = "green", alpha = 0.5, binwidth = 1) +
  labs(title = "protein A, treatment 2")

hist_pAt3 <- ggplot(pAt3,aes(ratio)) +
  geom_histogram(fill = "blue", alpha = 0.5, binwidth = 2) +
  labs(title = "protein A, treatment 3")

hist_pAt1 + hist_pAt2 + hist_pAt3 + violinplot_clean_pAt1 + ggtitle(label = "Protein A") + violinplot_clean_pAt2 + violinplot_clean_pAt3 + plot_layout(ncol=3,nrow=2) +
   plot_annotation(
  title = 'Treatment ratio for protein A',
  subtitle = 'filter at 470 nm excitation')# Plotting next to each other with patchwork


```

### 9A. Visualizing normality assumption with Q-Q plot


```{r}
set.seed(900)
 
par(mfrow = c(1,3))
qqnorm(pAt1$ratio, main = 'Protein A treatment 1')
qqline(pAt1$ratio, col = "steelblue", lwd = 2)

qqnorm(pAt2$ratio, main = 'Protein A treatment 2')
qqline(pAt2$ratio, col = "steelblue", lwd = 2)

qqnorm(pAt3$ratio, main = 'Protein A treatment 3')
qqline(pAt3$ratio, col = "steelblue", lwd = 2)

```

### 10A. Testing for normality of data

```{r}
# Testing for normality with the shapiro-Wilk test
shapiro.test(pAt1$ratio)
shapiro.test(pAt2$ratio)
shapiro.test(pAt3$ratio)
```

### 11A. Transformation of data with log() and testing normality

```{r}
# log() transformations
pAt1 <- pAt1  %>% 
              mutate(log_ratio = log(ratio))
shapiro.test(pAt1$log_ratio) # 0.01342

pAt2 <- pAt2  %>% 
              mutate(log_ratio = log(ratio))
shapiro.test(pAt2$log_ratio) # 0.02112

pAt3 <- pAt3  %>% 
              mutate(log_ratio = log(ratio))
shapiro.test(pAt3$log_ratio) # 0.1924
```

### 12A. Combine data

They all do not meet the normality criteria, even after transformation, log transformation being the closest to normality. Thus a non-parametric test can be used to look for differences in mean between groups.

Data will be combined in the long-format for hypothesis testing

```{r}
combined_pA <- rbind(pAt1,pAt2,pAt3)
```


### 13A. Checking the statistics of the cleaned data

```{r}
group_by(combined_pA, treatment) %>%
  summarise(
    count = n(),
    mean = mean(ratio),
    sd = sd(ratio),
    median = median(ratio),
    IQR = IQR(ratio)
  )
```

### 14A. Hypothesis testing: non-parametric Kruskal-Wallis ANOVA
Note that the difference is important in order to have a working protein.
The higher the ratio, then the protein indicates that the cell compartment is basic, and the other way around for acidic conditions.

H0: The means between groups are identical
H3: At least, the mean of one group is different

```{r}
kruskal_pA <- kruskal.test(log_ratio ~ treatment, data = combined_pA)
 # p < 2e-16*** (at least the mean of one group is different)
```


### 15A. Multiple pairwise-comparison between groups
According to the Kruskal-Walis test, there is a significant difference between groups with p < 0.05 ,however from the results, it is not possible to distinguish which groups differ.

```{r}
pairwise.wilcox.test(combined_pA$log_ratio, combined_pA$treatment,
                 p.adjust.method = "BH")
```


The test shows that the ratio is different for each treatment, which is a positive result, as they need to differ in order to be able to use as marker.

### 16A. Comparing (standardized) ratios, in proportion to treatment 2 as equivalent of 1.0

```{r}
means_pA <- combined_pA %>% # mean treatment 2 = 2.099387
            group_by(treatment) %>% 
            summarise(mean_ratio = mean(log_ratio), standardized_ratio = mean(log_ratio)/2.099387)

```


### 17A. Plotting the usual ratio and the standardized

```{r}
plot_pA_mean <- ggplot(means_pA, aes(x = treatment, y=mean_ratio)) +
      geom_point(size=2) +
      geom_text(aes(label = round(mean_ratio,2)), hjust = 0.5,  vjust = -1) +
      geom_line(aes(x=as.numeric(treatment),y=mean_ratio), color="steelblue") +
      scale_y_continuous(limits=c(2,3.5), breaks = seq(2,3.5,by=0.5))
      
plot_pA_std <- ggplot(means_pA,aes(x = treatment, y=standardized_ratio)) +
      geom_point(size=2) +
      geom_text(aes(label = round(standardized_ratio,2)), hjust = 0.5,  vjust = -1) +
      geom_line(aes(x=as.numeric(treatment),y=standardized_ratio), color="steelblue") +
      scale_y_continuous(limits=c(1,1.6), breaks = seq(1,1.6,by=0.2))

plot_pA_mean + ggtitle("Protein A") + plot_pA_std
```

------------------------------------------------------------------------------
## Fluorescence emission ratio at 470nm for Protein B with treatment 1, 2, 3.

### 1B. Organizing data and removing outliers
```{r}
pBt1 <- ratio_df %>%  # start with 171 observations
          group_by(treatment,protein,ratio) %>%
          filter(protein == "B" &
                 treatment == 1)

quantile1_pBt1 <- quantile(pBt1$ratio,0.25) # 16.965 
quantile3_pBt1 <- quantile(pBt1$ratio,0.75) # 21.615
IQR_pBt1 <- quantile3_pBt1 - quantile1_pBt1 # 4.65

pBt1 <- pBt1 %>% 
          mutate(is_outlier = ratio < quantile1_pBt1 - (IQR_pBt1*1.5) |
                   ratio >  quantile3_pBt1 + (IQR_pBt1*1.5)) %>% 
          filter(!is_outlier) # remains 163 observations, 8 outliers removed

#--------------------------------------------------------
pBt2 <- ratio_df %>%  # start with 138 observations
          group_by(treatment,protein,ratio) %>%
          filter(protein == "B" &
                 treatment == 2)

quantile1_pBt2 <- quantile(pBt2$ratio,0.25) #20.33
quantile3_pBt2 <- quantile(pBt2$ratio,0.75) #27.3025 
IQR_pBt2 <- quantile3_pBt2 - quantile1_pBt2 # 6.96

pBt2 <- pBt2 %>% 
          mutate(is_outlier = ratio < quantile1_pBt2 - (IQR_pBt2*1.5) |
                   ratio >  quantile3_pBt2 + (IQR_pBt2*1.5)) %>% 
          filter(!is_outlier)# remains 137 observations, 1 outlier removed

#---------------------------------------------------------
pBt3 <- ratio_df %>%  # start with 130 observations
          group_by(treatment,protein,ratio) %>%
          filter(protein == "B" &
                 treatment == 3)

quantile1_pBt3 <- quantile(pBt3$ratio,0.25) # 29.0075
quantile3_pBt3 <- quantile(pBt3$ratio,0.75) # 37.18
IQR_pBt3 <- quantile3_pBt3 - quantile1_pBt3 # 8.1725

pBt3 <- pBt3 %>% 
          mutate(is_outlier = ratio < quantile1_pBt3 - (IQR_pBt3*1.5) |
                   ratio >  quantile3_pBt3 + (IQR_pBt3*1.5)) %>% 
          filter(!is_outlier)# remains 124 observations, 6 outliers removed
```



### 2B. Plotting ratios without outliers

```{r}
violinplot_clean_pBt1 <- pBt1 %>% 
  ggplot(aes(x = "treatment 1", y = ratio)) +
  geom_violin(alpha=0.5,fill = "red") + geom_boxplot(outlier.color="blue", alpha=0.5) +
  xlab("")

violinplot_clean_pBt2 <- pBt2 %>% 
  ggplot(aes(x = "treatment 2", y = ratio)) +
  geom_violin(alpha=0.5,fill = "green") + geom_boxplot(outlier.color="red", alpha=0.5) +
  xlab("")

violinplot_clean_pBt3 <- pBt3 %>% 
  ggplot(aes(x = "treatment 3", y = ratio)) +
  geom_violin(alpha=0.5,fill = "blue") + geom_boxplot(outlier.color="red", alpha=0.5) +
  xlab("")
```

### 3B. Viewing distribution in a histogram and combining all graphs

```{r}
set.seed(5)
hist_pBt1 <- ggplot(pBt1,aes(ratio)) +
  geom_histogram(fill = "red", alpha = 0.5, binwidth = 2) +
  labs(title = "Protein B at treatment 1")

hist_pBt2 <- ggplot(pBt2,aes(ratio)) +
  geom_histogram(fill = "green", alpha = 0.5, binwidth = 2) +
  labs(title = "Protein B at treatment 2")

hist_pBt3 <- ggplot(pBt3,aes(ratio)) +
  geom_histogram(fill = "blue", alpha = 0.5, binwidth = 2) +
  labs(title = "Protein B at treatment 3")

hist_pBt1 + hist_pBt2 + hist_pBt3 + violinplot_clean_pBt1 + ggtitle(label = "Protein B") + violinplot_clean_pBt2 + violinplot_clean_pBt3 + plot_layout(ncol=3,nrow=2) +
   plot_annotation(
  title = 'Treatment ratio for Protein B',
  subtitle = 'GFP filter at 470 nm excitation')# Plotting next to each other with patchwork

```

### 4B. Visualizing normality assumption with Q-Q plot

```{r}
set.seed(900)
 
par(mfrow = c(1,3))
qqnorm(pBt1$ratio, main = 'Protein B treatment 1')
qqline(pBt1$ratio, col = "steelblue", lwd = 2)

qqnorm(pBt2$ratio, main = 'Protein B treatment 2')
qqline(pBt2$ratio, col = "steelblue", lwd = 2)

qqnorm(pBt3$ratio, main = 'Protein B treatment 3')
qqline(pBt3$ratio, col = "steelblue", lwd = 2)
```


### 5B. Testing normality with Shapiro-Wilk test.

```{r}
# Testing for normality with the shapiro-Wilk test
shapiro.test(pBt1$ratio) # p = 0.9338
shapiro.test(pBt2$ratio) # p = 0.2168
shapiro.test(pBt3$ratio) # p = 0.777
```


They all meet the normality criteria, thus a parametric test (ANOVA) can be used to look for differences in mean between groups.

### 6B. Data combined in the long-format for hypothesis testing.

```{r}
combined_pB <- rbind(pBt1,pBt2,pBt3)
```

### 7B. Hypothesis testing.

H0: The means between groups are identical
H3: At least, the mean of one group is different

```{r}
anova_combined_pB <- aov(ratio~treatment, combined_pB)
summary.aov(anova_combined_pB) # p < 2e-16*** (at least the mean of one group is different)
model.tables(anova_combined_pB, "means") # table of means
```



There is a significant difference between groups, but which groups exactly differ?

### 8B. Multiple pair-wise comparison

```{r}
pairwise.wilcox.test(combined_pB$ratio, combined_pB$treatment,
                 p.adjust.method = "BH")
```


All groups differ significantly. 

### 9B. Comparing (standardized) ratios

```{r}
means_pB <- combined_pB %>% 
            group_by(treatment) %>% 
            summarise(mean_ratio = mean(ratio), standardized_ratio = mean(ratio)/23.91153)

```

```{r}

plot_pB_mean <- ggplot(means_pB,aes(x = treatment, y=mean_ratio)) +
      geom_point(size=2) +
      geom_text(aes(label = round(mean_ratio,2)), hjust = 0.5,  vjust = -1) +
      geom_line(aes(x=as.numeric(treatment),y=mean_ratio), color="steelblue") +
      scale_y_continuous(limits=c(15,35), breaks = seq(15,35,by=5))

plot_pB_std <- ggplot(means_pB,aes(x = treatment, y=standardized_ratio)) +
      geom_point(size=2) +
      geom_text(aes(label = round(standardized_ratio,2)), hjust = 0.5,  vjust = -1) +
      geom_line(aes(x=as.numeric(treatment),y=standardized_ratio), color="steelblue") +
      scale_y_continuous(limits=c(0.8,1.4), breaks = seq(0.8,1.4,by=0.2))
plot_pB_mean + ggtitle("Protein B") + plot_pB_std
```








